---
title: "Getting started with causalbic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with causalbic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(causalbic)
```

## The model

`causalbic` is for a proportional outcome `y` in `[0, 1]` with a point
mass at a boundary (zero, one, or both), and a binary treatment that may
act two ways: by moving the probability of landing on the boundary, or by
shifting the mean of the cases that do not. `simulate_bic()` draws from
such a model with a known, closed-form effect and its two components, so
the rest of this vignette can check the estimator against ground truth.

```{r}
set.seed(1)
dat <- simulate_bic(600)
head(dat)
attr(dat, "truth")
```

## Fitting and the effect decomposition

`bic_fit()` fits the inflated beta regression (`gamlss.dist::BEZI` for one
boundary, `gamlss.dist::BEINF` for two, chosen automatically by whether
`tau.formula` is supplied). `bic_gcomp()` then estimates the average
treatment effect and decomposes it exactly into a discrete-component part
(through the boundary probability) and a continuous-component part
(through the conditional mean):

```{r}
fit <- bic_fit(y ~ T + W2, nu.formula = ~ T + W1, data = dat)
g <- bic_gcomp(fit, treat = "T", data = dat)
g
g["tau_alpha"] + g["tau_mu"] - g["tau"]  # exact identity
```

The conditional effect at specific covariate profiles uses the same
decomposition, `bic_cate()`:

```{r}
profiles <- data.frame(W1 = c(-1, 0, 1), W2 = c(-1, 0, 1))
bic_cate(fit, "T", profiles)
```

## Diagnostics: curvature, influence and sensitivity

`bic_curvature()` builds the causal conformal normal curvature in the
treatment direction and splits it, additively and exactly, into the same
two components:

```{r}
cv <- bic_curvature(fit, "T", dat)
c(BdT = cv$BdT, discrete = cv$BdT_discrete, continuous = cv$BdT_continuous)
```

`bic_icim()` reads the per-unit diagonal of that curvature off to flag
disproportionately influential observations, and classify them by which
component they act through:

```{r}
icim <- bic_icim(cv)
sum(icim$influential)
table(icim$type)
```

`bic_sensitivity()` bounds how much an unmeasured confounder of a given
strength could move the estimate, and solves for the smallest confounding
strength `Gamma_star` that would be needed to erase the effect. This bound
can be, and on some data sets is, uninformative (`Gamma_star` close to 1
or `NA`): it says what it says honestly rather than manufacturing
reassurance.

```{r}
sens <- bic_sensitivity(cv, fit, "T", dat, tau_hat = g["tau"])
c(L1 = sens$L1, Gamma_star = sens$Gamma_star)
```

## Reference

Ospina, R. (2026). Causal inference for proportional outcomes via
likelihood displacement in inflated beta regression. Submitted to *The
Annals of Applied Statistics*.
