---
title: "Reproducing the simulation study (Ospina, 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the simulation study (Ospina, 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette reproduces the simulation study of Section 7 of Ospina
(2026), using exactly the data-generating process and parameters of the
manuscript's `paper2_simulation.R` (the defaults of [simulate_bic()]
are that same process). It runs end to end from package functions alone,
with no external data.

```{r setup}
library(causalbic)
```

## What `tau_alpha` and `tau_mu` mean

Before looking at numbers: `tau` (the average treatment effect) is the
total change in the mean response caused by treatment. `tau_alpha` is the
part of that change coming from the treatment moving units onto or off
the boundary (here, the probability of landing at `y = 0`); `tau_mu` is
the part coming from the treatment shifting the conditional mean among
units that remain in the interior. They always add up to `tau` exactly
(see below), but they can, and in this design do, point in opposite
directions, so a reader who only sees `tau` cannot tell whether the
treatment acted on the boundary mechanism, the interior mechanism, both
in agreement, or both in opposition with one dominating.

## Population truth

The manuscript reports, from a population of $2\times10^6$ draws under
`gamma = c(-1.5, 0.3, 1.5)`, `beta = c(-1.0, 0.5, 1.2)`, `phi = 4`,
`rho = 0.4`: $\tau = 0.0463$, $\tau_\alpha = -0.0876$,
$\tau_\mu = 0.1338$. A large simulated population reproduces this closely:

```{r}
set.seed(1)
pop <- simulate_bic(2e5, seed = 1)
unlist(attr(pop, "truth"))
```

Reading this substantively: the treatment actually *increases* the
boundary probability ($\tau_\alpha < 0$ here means a lower mean
contribution from a higher chance of landing at the boundary, since the
boundary is $y=0$) while it *raises* the interior mean enough to overcome
that pull ($\tau_\mu>0$, and larger in magnitude than $\tau_\alpha$).
The two components have opposite signs and are each about three times
larger than their difference, $\tau$: the average effect alone would
describe this as a modest, roughly homogeneous effect and conceal that
two mechanisms of comparable size are pulling in opposite directions.

## Estimator behavior at finite samples

The manuscript's Table 2 reports bias, RMSE and coverage at
$n \in \{100, 200, 500\}$ over $R = 200$ replications. The loop below
does the same at a smaller `R`, for vignette build time; set `R <- 200`
to reproduce Table 2 exactly (seeds match `paper2_simulation.R`).

```{r}
run_once <- function(n, seed) {
  dat <- simulate_bic(n, seed = seed)
  fit <- bic_fit(y ~ T + W2, nu.formula = ~ T + W1, data = dat)
  bic_gcomp(fit, "T", dat)
}

R <- 20  ## paper uses R = 200
ns <- c(100, 200, 500)
truth <- unlist(attr(simulate_bic(10, seed = 1), "truth"))

results <- do.call(rbind, lapply(ns, function(n) {
  ests <- t(vapply(seq_len(R), function(r) run_once(n, seed = r), numeric(3)))
  bias <- colMeans(ests) - truth[c("tau", "tau_alpha", "tau_mu")]
  rmse <- sqrt(colMeans(sweep(ests, 2, truth[c("tau", "tau_alpha", "tau_mu")])^2))
  data.frame(n = n, bias_tau = bias["tau"], rmse_tau = rmse["tau"])
}))
rownames(results) <- NULL
results
```

RMSE should decline roughly as $n^{-1/2}$ as $n$ grows, and bias should
stay an order of magnitude below the effect size ($\tau = 0.046$) at
every $n$, matching Table 2 of the manuscript. In practical terms, this
table is the check that the estimator earns the right to be trusted on
real data before Section 8 uses it there: a decomposition that looked
exact algebraically but recovered the wrong $\tau_\alpha$/$\tau_\mu$
split in finite samples would be a mathematical curiosity, not an
estimator.

## The decomposition is an exact identity

Regardless of `R` or `n`, the decomposition holds to floating-point
precision on every single fit, not just on average:

```{r}
dat <- simulate_bic(300, seed = 99)
fit <- bic_fit(y ~ T + W2, nu.formula = ~ T + W1, data = dat)
g <- bic_gcomp(fit, "T", dat)
g["tau"] - (g["tau_alpha"] + g["tau_mu"])
```

This is not a numerical coincidence to be checked once and trusted: it
follows from the inflated beta likelihood being separable (its Fisher
information is block diagonal between the boundary-probability and mean
submodels), so `tau_alpha` and `tau_mu` are not two separate estimates
that happen to sum correctly, they are the same single fitted model read
two ways. That is also why misspecifying the boundary submodel does not
contaminate the mean submodel's parameters (though it does bias the
*weight* `tau_mu` is averaged with, since that weight is a boundary
probability); see Section 4 of the manuscript.

## Figures

`causalbic` also ships the plotting functions behind two of the
manuscript's figures, so they can be regenerated from the package alone,
not just their underlying numbers. Both are fully self-contained (no
external data): [plot_concept()] draws the synthetic two-panel
illustration of the discrete/continuous decomposition and the
likelihood-displacement curve behind the causal curvature (Figure 1);
[plot_cate()] simulates one dataset from the same process as above, fits
it, and plots the true and estimated conditional average treatment effect
alongside its two components (Figure 2).

```{r, fig.width = 9.2, fig.height = 4.6}
par(mfrow = c(1, 2))
plot_concept()
```

```{r, fig.width = 9.2, fig.height = 4.2}
par(mfrow = c(1, 2))
invisible(plot_cate())
```

(A third figure, the application's goodness-of-fit diagnostics, needs a
fitted model on real data and is shown in `vignette("paper-application")`
via [plot_diagnostics()].)

## Reference

Ospina, R. (2026). Causal inference for proportional outcomes via
likelihood displacement in inflated beta regression. Submitted to *The
Annals of Applied Statistics*.
